---
title: "Statisitcal Programming - 2020 - Project 1"
output: rmarkdown::html_document
author: Megan Richards
---

Due on Monday, November 2nd by 5:00 pm (17:00), UK local time.

Before you start be sure to read through *all* of the rules and instructions in the `README.md`.

<br/>

### Data

```{r}
library(tidyverse)
library(RColorBrewer)
library(gifski)
library(gganimate)
f1 = readRDS(file="data/f1.rds")
```


<br/>

### Task 1 - Tidy the data

#### Write up

To begin, we convert the f1 data into a tibble which is a data frame chosen for its useful properties when it comes to working in the tidyverse. Creating a tibble results in a list column which we can then manipulate and pull information from. Using the 'View' function in the console allows us to see that there are 14 entries in this data structure, each with a list of MRData containing the information we are interested in. 

Therefore, the first step is to use the unnest_wider function such that we progress into the MRData level of the data structure. Next, we notice that as we progress through this data structure, we come across more and more redundant data. Thus, we begin to use the hoist function to extract only the information we are interested in. For example, we hoist only the information from the RaceTable within the MRData column and then within this newly created column RaceTable, we hoist only the information contained within Races. In order to discard irrelevant information, we use the select function to remove the columns once they have had the relevant information extracted. This is standard practice and is repeated after every application of hoist in this task. 

At this point, we have obtained a list column with each list containing two entries. As we want the information contained within both entries, we use the unnest_longer function to expand this list. We then hoist again to obtain the columns race_name, round, date and results. Finally, we have obtained some columns which should appear in the final tidy table - this is promising! However, we note that the type of these columns is different from the type specified in the task. Therefore, we can mutate these columns to coerce their entries into the desired types. Now, we can leave the first 3 columns and focus on getting the remaining information from the column 'results'. 

Since the results for each driver are obtained within this list, we want to apply unnest_longer to this column. In the race_name column, we now get every race repeated. This is what we want, since we eventually want to obtain rows where each represents the result of a driver for a particular race. 

Now, hoist the information 'Driver', 'Constructor', 'positionText' and 'points' from the results column. We choose to hoist the 'positionText' entries rather than the entries in 'position'. This is because the value of the position attribute in the Result element is always an integer, giving the finishing order of all drivers. However, the value of the positionText attribute is either an integer (finishing position), “R” (retired), “D” (disqualified), “E” (excluded), “W” (withdrawn), “F” (failed to qualify) or “N” (not classified). This is important, since we want to replace the position of the driver with NA if they did not finish for any reason. We do this step now, using the mutate function to replace the position value with NA if it is one of the letters above, but leaving it as the finishing position otherwise. Then, we mutate the columns points and position to have integer types. 

Our table now looks close to the desired result. The next step is to hoist the names of each driver from the driver information and the names of the constructors from the constructor information. Since we want the full names of the drivers, we hoist both the first names and surnames. Then, we use the mutate function along with the paste function to create a new column combining the first and surnames of the drivers into a single entry of their full name. We can then use the select function to remove the columns of first and surname. Finally, we use the relocate function to reorder the columns to match the requirements of the task. 

This data is now tidy, since each variable has its own column, each observation has its own row and each value has its own cell. Therefore, we can use the head function to print out the first 10 rows of this data frame to finish the instructions of the task. The kable function is used so the table is displayed nicely in the markdown file. 


#### Code

```{r task1}
f1_tidy = f1 %>%
  tibble::tibble(f1 = .) %>%
  unnest_wider(f1)%>%
  hoist(
    MRData,
    table = c("RaceTable")
  )%>%
  select(-MRData)%>%
  hoist(
    table,
    races = c("Races")
  )%>%
  select(-table)%>%
  unnest_longer(races)%>%
  hoist(
    races,
    race_name = c("raceName"),
    round = c("round"),
    date = c("date"),
    results = c("Results")
  )%>%
  mutate(date = as.Date(date))%>%   # change type of column
  mutate(round = as.integer(round))%>%
  select(-races)%>%
  unnest_longer(results)%>%
  hoist(
    results,
    driver_info = c("Driver"),
    constructor_info = c("Constructor"),
    position = c("positionText"),
    points = c("points")
  )%>%
  mutate(position = na_if(position, "R"))%>%  # replace w/ NA if DNF
  mutate(position = na_if(position, "W"))%>%
  mutate(position = na_if(position, "D"))%>%
  mutate(position = na_if(position, "E"))%>%
  mutate(position = na_if(position, "F"))%>%
  mutate(position = na_if(position, "N"))%>%
  mutate(position = as.integer(position))%>%
  mutate(points = as.integer(points))%>%
  select(-results)%>%
  hoist(
    driver_info,
    first_name = c("givenName"),
    surname = c("familyName")
  )%>%
  select(-driver_info)%>%
  hoist(
    constructor_info,
    constructor = c("name")
  )%>%
  select(-constructor_info)%>%
  mutate(driver = paste(first_name, surname, sep=" "))%>% # combine drivers'                                                             first and surnames
  select(-first_name, -surname)%>%
  relocate(race_name, round, date, driver)  # rearrange order of table

knitr::kable(head(f1_tidy, 10)) # print first 10 rows
```

<br/>


### Task 2 - Drivers' Championship

#### Write up

We begin by abbreviation of the race names such that the table can fit onto a knitted document. Each race name is unique and easily identifiable. We progress by selecting only the rows of the data frame that are relevant to this task. This could be done by selecting relevant rows or deselecting redundant rows - we chose the former since the computational effort for each option was similar and I believe this option is more obvious. 

Next, we create a "grouped" copy of the table such that we can manipulate the results for each driver individually. We then add a column using the mutate function to sum over the points achieved by each individual driver across all races. Once we have this column of overall points, we no longer need to manipulate the points achieved in each individual race and therefore we remove this column from the data frame with the negation of the select function. 

Since we want to achieve a table with a row for each driver containing their positions in each race, we now want to pivot the data frame so that the race names are expanded into individual columns. We achieve this by the pivot_wider function, obtaining the column names from race_name and the values displayed from the corresponding position of the driver in that particular race.

Finally, we want to change the order of the data frame such that it is now sorted in descending order of points total, rather than in the order the drivers finished in the first race only. We achieve this through application of the arrange function on the overall points column along with the function desc since we want the highest number of points at the top of the table. To conclude, we reorder the columns such that the overall points column is moved to the final column of the table. 

Note that we have assigned this data frame a name 'f1_task2' so that we can reference it in later tasks. Therefore, we must include a line outside of the pipe to print the table. This is the final line in the chunk. 

#### Code

```{r task2, warning = FALSE}

# abbreviate race names to make table narrower
f1_tidy[f1_tidy$race_name == "Australian Grand Prix", 1] <- "AUS"
f1_tidy[f1_tidy$race_name == "Bahrain Grand Prix", 1] <- "BHR"
f1_tidy[f1_tidy$race_name == "Chinese Grand Prix", 1] <- "CHN"
f1_tidy[f1_tidy$race_name == "Azerbaijan Grand Prix", 1] <- "AZE"
f1_tidy[f1_tidy$race_name == "Spanish Grand Prix", 1] <- "ESP"
f1_tidy[f1_tidy$race_name == "Monaco Grand Prix", 1] <- "MON"
f1_tidy[f1_tidy$race_name == "Canadian Grand Prix", 1] <- "CAN"
f1_tidy[f1_tidy$race_name == "French Grand Prix", 1] <- "FRA"
f1_tidy[f1_tidy$race_name == "Austrian Grand Prix", 1] <- "AUT"
f1_tidy[f1_tidy$race_name == "British Grand Prix", 1] <- "GBR"
f1_tidy[f1_tidy$race_name == "German Grand Prix", 1] <- "GER"
f1_tidy[f1_tidy$race_name == "Hungarian Grand Prix", 1] <- "HUN"
f1_tidy[f1_tidy$race_name == "Belgian Grand Prix", 1] <- "BEL"
f1_tidy[f1_tidy$race_name == "Italian Grand Prix", 1] <- "ITA"
f1_tidy[f1_tidy$race_name == "Singapore Grand Prix", 1] <- "SIN"
f1_tidy[f1_tidy$race_name == "Russian Grand Prix", 1] <- "RUS"
f1_tidy[f1_tidy$race_name == "Japanese Grand Prix", 1] <- "JPN"
f1_tidy[f1_tidy$race_name == "Mexican Grand Prix", 1] <- "MEX"
f1_tidy[f1_tidy$race_name == "United States Grand Prix", 1] <- "USA"
f1_tidy[f1_tidy$race_name == "Brazilian Grand Prix", 1] <- "BRA"
f1_tidy[f1_tidy$race_name == "Abu Dhabi Grand Prix", 1] <- "ABU"

f1_task2 = f1_tidy%>%
  select(driver, race_name, position, points)%>%
  group_by(driver)%>% # want to work with each driver separately
  mutate(overall_points = sum(points))%>% # total all points in new column
  select(-points) %>%
  pivot_wider(
    names_from = race_name, 
    values_from = position
  )%>%
  arrange(desc(overall_points))%>%  # order the table
  relocate(c("overall_points"), .after = last_col())  # move overall points to                                                         final column

knitr::kable(f1_task2) # print table
```
<br/>

### Task 3 - Cumulative Constructors

#### Write up

As usual, we begin by selecting only the columns of the tidy data frame that will be required for this task. Specifically, these are the constructor, race_ name and points columns. 

Next, we implement the pivot_wider function to find the points achieved by each constructor in each race. This forms a table with a column for each constructor and a row for each column. Since each constructor has two drivers, we obtain a list in each entry of this data frame. These lists contain two entries - the points obtained by each of the constructor's drivers in the particular race. Hence, we want to sum the entries in these lists to obtain one value; namely the total number of points achieved by the constructor for that race. To achieve this, we implement the conditional variant of the mutate function 'mutate_if'. We use this function with the predicate 'is.list' such that we only mutate the columns that are of list type (i.e. all columns except the race name). Within these columns, we use the map_dbl function from purrr to apply a function to each element of the lists. We use '.' to refer to the previously selected columns only, and apply the sum function element-wise to each list. Note that we use map_dbl to return a result of double type. 

We are getting there - we have obtained a table with values corresponding to the total points achieved by each constructor in each race. However, we want the cumulative points achieved. Again, we use the mutate_if function to select only the columns of numeric type since we cannot (and do not want to) take a cumulative sum over the race names. To these columns, we apply the cumsum function to return a cumulative points tally as we progress through the races (i.e. down the column) for each constructor.

Now, we have achieved the desired result but in the wrong format. We want the constructors in the first column such that they correspond to the rows of the data frame; and we want the columns of the data frame to be the race names. Since there is no easy way to transpose a data frame in R, we will pivot the table twice. 

We first pivot_longer to obtain a table with the columns race_name, constructor and (cumulative) points. Since we do not want to touch the race_name column, we refer to the data frame to be pivoted with !race_name. Then, we pivot_wider with the column names coming from race_name and the values coming from the (cumulative) points as before. In this way, we achieve a data frame of the desired format. Finally, we want to sort the data frame in order of descending points achieved by each constructor at the end of all the races. To do this, we use the arrange function on a subset of the data frame corresponding to the final column. Through application of the rev function, we reorder the columns such that subsetting with the index 1 corresponds to the final column, rather than the first column. This makes the code more robust than just subsetting the data frame with index 22, for example we could input data for another race and the code would still work as intended. The minus sign corresponds to descending ordering as desired.

Note, again we have assigned this data frame a name 'f1_task3' so that we can reference it in later tasks. Therefore, we must include a line outside of the pipe to print the table. This is the final line in the chunk. 

#### Code


```{r task3, warning= FALSE}
f1_task3 = f1_tidy%>%
  select(constructor, race_name, points)%>%
  pivot_wider(
    names_from = constructor, 
    values_from = points
  )%>%
  mutate_if(    
    is.list,        # select only the columns containing lists
    ~ purrr::map_dbl(., sum)     # sum over these list entries
  )%>%
  mutate_if(
    is.numeric,     # select all columns except race name
    cumsum)%>%      # give a cumulative sum
  pivot_longer(     # begin process of switching rows and columns 
    !race_name, 
    names_to = "constructor", 
    values_to = "points")%>%
  pivot_wider(
    names_from = race_name, 
    values_from = points
  )%>%
  arrange(-rev(.)[1]) # minus sign for descending, rev for last col. 

knitr::kable(f1_task3)
```

<br/>

### Task 4 - Visualization

#### Write up

I began this question by thinking about the information I wanted to portray through my visualizations. I decided to plot bar graphs as overviews of the  performances of both drivers and teams over the course of the 2019 F1 season. This was so that I could see all the information necessary in a way that would allow me to then decide what were the important focal points. 

From the bar chart for the drivers' performances, I noticed that there was a wide range of results; in particular, the top 5 drivers could be seen to be ahead of the rest of the competition since the beginning of the season. Since the range of performance was so vast, I noticed that the scale on the bar chart was difficult to read at places. Thus, I decided to utilize faceting to display smaller plots showing each of the drivers individual performances. I adjusted the scale such that it was no longer shared between all plots, but free to adjust to the individual drivers performance. This allowed us to see the performance of each driver in more detail - for example, we see that Lewis Hamilton had a steady season, accumulating a similar number of points consistently from each race whereas Kevin Magnussen performed far better in some races than others. These plots further emphasized the dominance of the top 5 drivers. Therefore, I decided to explore the performances of only these drivers through a line graph plotting each of their cumulative points tallies as the season progressed. This made the results more clear since a smaller axis range could be used by eliminating the results of the rest of the drivers. Further, I decided to animate this graph such that the reader could then see the times in the season where certain point milestones were reached. This added further insight to the standings throughout the season, and showed the dominance of Lewis Hamilton as he was the first to reach the majority of point milestones.

I then moved onto the constructors' performances. Again, I began by plotting a bar graph to show as much information as possible. I noted that in this plot, it was the top 3 constructors that dominated the season such that the scale was difficult to read for the remaining constructors. Therefore, I decided to plot a line graph showing the performances of constructors outside of the top three throughout the season. This was an improvement to the story told through visualization, since we can now see in detail the changes in rankings between these drivers as the season progressed. For example, it was difficult to see in the bar graph whether Williams obtained any points, or if they remained on zero throughout the season. The line graph clearly shows they obtained points in the German Grand Prix. I will now explain the method to achieve these graphs. 

As usual, I began by selecting the rows of the tidy data frame that would be relevant to this task. I then grouped by driver such that I could manipulate the drivers' individual results. Then, I noted that the rainbow-like default colors used in the plot made it difficult to distinguish between races since the colors seemed to merge into each other. Therefore, I counted the number of distinct race names (such that each could be assigned a unique color) and changed the color palette to make the distinct races more obvious. In addition, I added an outline to the bars to further emphasize the distinct nature of each race. I ensured that the bar graph was ordered by points achieved at the end of the season with the driver in poll position at the top of the graph. Finally, I added a legend, labeled the axes and titled the graph. 

Next, in order to plot my second graph I had to manually override R such that the races were displayed in chronological order, rather than alphabetically. To do this, I created an ordered list called "countries" corresponding to the 3 letter abbreviations of race names in chronological order; this was then used to set the levels. I could then plot the graphs to show drivers' individual performances utilising the technique of faceting. I chose to set the number of columns to 2 such that each individual plot was a good size. 

In order to select the top 5 drivers, I chose to refer to them by their position in the table rather than by name. This ensures the code is more robust to changes, such that if the top 5 were to change after additional races the code would continue to work. Again, we needed to manually override the order of the x axis as before. The default colors were acceptable, but I decided to increase the increments displayed on the y axis. I labeled the axes, provided a legend and added a meaningful title to explain one of the reasons I decided to focus on a smaller number of drivers - so that the changes in positions are easier to see as intersections of the lines. 

Similar methods were followed for the constructor visualizations. The code for this bar chart is identical to the previous, with the driver variable changed now to constructor name. Again, I used position within the table to select the constructors outside of the top 3. I used the negation of the top 3 constructors rather than selecting those constructors in positions 4 to 10. This is so that, if more constructors are added, the graph will continue to show what it is supposed to as the code is robust.

#### Code

```{r task4, fig.height=14, fig.width=14, warning = FALSE}
driver_points = f1_tidy%>%
  select(driver, race_name, points)%>%
  group_by(driver) 

# get a unique color for each race
colourCount = length(unique(driver_points$race_name)) 

# change color palette to make more obvious
getPalette = colorRampPalette(brewer.pal(9, "Set1"))


ggplot(
  driver_points, 
  aes(
    y = reorder(driver, points),
    x = points
    )
  )+
  geom_col(aes(fill = race_name), color = "black", width = 0.7)+  #add outline theme(legend.position="bottom") + 
  labs(
    title = "Points achieved cumulatively by drivers throughout the season",
    subtitle = paste("Bars stacked such that points achieved in each race are visible"),
    x = "Cumulative Points",
    y = "Driver")+
  scale_fill_manual(values = getPalette(colourCount))+
  guides(fill=guide_legend(title="Race Name"))

# get a cumulative sum of points
driver_cum_points = driver_points%>%
  mutate(points = cumsum(points))

# ensure ordering of races is chronological
countries = unique(driver_cum_points[[2]])
driver_cum_points$race_name <- factor(driver_cum_points$race_name, levels = countries)

ggplot(
  driver_cum_points,
  aes(
    x = race_name, 
    y = points,
    color = driver
    )
  ) +
  geom_point() +
  facet_wrap(~driver, scales = "free", ncol = 2) + # change scales of plots
guides(color=FALSE)+ 
  labs(
    title = "Points achieved cumulatively by drivers throughout the season",
    subtitle = paste("Split to show each drivers individual performance in each race"),
    x = "Race name",
    y = "Cumulative Points")


# obtain only the top 5 drivers' cumulative points
# f1_task2[[1]] gives us list of all drivers
driver_cum_points_top5 = driver_cum_points%>%
  filter(driver == f1_task2[[1]][1]| driver ==  f1_task2[[1]][2] |driver == f1_task2[[1]][3] |driver == f1_task2[[1]][4]| driver == f1_task2[[1]][5])

# ensure chronological ordering
driver_cum_points_top5$race_name <- factor(driver_cum_points_top5$race_name, levels = countries)

ggplot(
  driver_cum_points_top5, # plot only the top 5 drivers
  aes(
    x = race_name, 
    y = points,
    group = driver,
    color = driver
    )
  )+
  geom_line() +
  geom_point() + 
  gganimate::transition_reveal(points)+ # add an animation
  labs(
    title = "Driver in poll position changes from beginning to end of season",
    subtitle = paste("Animation to show milestone point tallies reached by top 5 drivers"),
    x = "Race name",
    y = "Cumulative Points")+
  scale_y_continuous(breaks = round(seq(min(driver_cum_points_top5$points), max(driver_cum_points_top5$points), by = 20),1)) # change increment of scale




# repeat for constructor

cons_points = f1_tidy%>%
  select(constructor, race_name, points)%>%
  group_by(constructor) 
  
ggplot(
  cons_points, 
  aes(
    y = reorder(constructor, points),
    x = points
    )
  )+
  geom_col(aes(fill = race_name), color = "black", width = 0.7)+ theme(legend.position="bottom") + 
  labs(
    title = "Points achieved cumulatively by constructors throughout the season",
    subtitle = paste("Bars stacked such that points achieved in each race are visible"),
    x = "Cumulative Points",
    y = "Constructor")+
  scale_fill_manual(values = getPalette(colourCount))+
  guides(fill=guide_legend(title="Race Name"))

# get cumulative points
cons_cum_points = cons_points%>%
  mutate(points = cumsum(points))

# ensure chronological ordering
cons_cum_points$race_name <- factor(cons_cum_points$race_name, levels = countries)


#The top three constructors are way in front of the others
#focus on others 

# count total constructors  - more robust
total_cons = length(f1_task3[[1]]) 
# check we have selected correct constructors
bottom_cons = f1_task3[[1]][4:total_cons]

# select all but top 3
cons_cum_points_bottom = cons_cum_points%>%
  filter(!(constructor == f1_task3[[1]][1]) & !(constructor == f1_task3[[1]][2]) & !(constructor == f1_task3[[1]][3]))

ggplot(
  cons_cum_points_bottom, 
  aes(
    x = race_name, 
    y = points,
    group = constructor,
    color = constructor
    )
  )+
  geom_line() +
  geom_point() + 
  labs(
    title = "Focus on cumulative points of constructors outside the top 3",
    x = "Race Name",
    y = "Cumulative Points")+
  scale_y_continuous(breaks = round(seq(min(cons_cum_points_bottom$points), max(cons_cum_points_bottom$points), by = 10),1)) # change y scale increments
```

<br/>

